Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-12 paper.tex
Rephrase

The present work is but the beginning of our research program. We
shall focus \emph{exclusively} on the semantics aspects, only briefly
touching upon implementation aspects (which remain the raison d'être
of the project!). Our experience designing the Usuba~\citep{mercadier:PhD}
programming language and implementing its compiler gives us some
confidence that a syntax can indeed be tailored to suit this
semantics.

We nonetheless could not resist the temptation of developing the
semantics in the Coq theorem prover, as executable programs. After
all, we are interested in combinational circuits: if there has ever
been a time where we do not have to worry about termination, this is
it! So we have implemented cryptographic primitive in pure Gallina.
This is nothing to worry about: after all, more ambitious colleagues
went from there down to deeply-embedded assembly
code~\citep{chlipala:coq-crypto}. Besides the ability to prove the
correctness and, even more usefully, test our code, this also enables
us to easily prototype a fixslicing compiler, using Coq's
\texttt{autorewrite} tactics to emulate a transformation-based simplification
engine~\citep{peyton-jones:haskell-optim} (part of an imaginary optimizing
compiler).

@@

This article grew out of the Bachelor's research project of the first
author and the Master thesis of the second author. The former worked
on the specification, bitslicing and fixslicing of the Skinny
cipher~\citep{beierle:skinny, adomnicai:fixslicing-AES-like}. The
latter worked on the specification, bitslicing and fixslicing of the
Gift cipher~\citep{subhadeep:gift, adominicai:fixslicing-gift}. Both
ciphers are provided in the accompanying source
code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
we shall focus exclusively on Skinny as our running example here.

Note that none of the programs below were easy to write in the first
place: it took weeks of careful study to weed out the essential
complexity from the accidental mismanagement of array indices. If the
code seem somewhat trivial, one should bear in mind that this
simplicity was hard-won. All the more reason to, another day, design a
programming language, so as to automatically inhabit this semantics,
and write its compiler, so as to automatically optimize code following
our theorems.

>>>>

The present work is but the beginning of our research program. We
shall focus \emph{exclusively} on the semantics aspects, only briefly
touching upon implementation aspects (which remain the raison d'être
of the project!). Our experience designing the Usuba~\citep{mercadier:PhD}
programming language and implementing its compiler gives us some
confidence that a syntax can indeed be tailored to dress up this
semantics.

We nonetheless could not resist the temptation of developing the
semantics in the Coq theorem prover, as executable programs. After
all, we are interested in combinational circuits: if there has ever
been a time where we do not have to worry about termination, this is
it! So we have implemented cryptographic primitive in pure Gallina.
This is nothing to worry about: after all, more ambitious colleagues
went from there down to deeply-embedded assembly
code~\citep{chlipala:coq-crypto}. Besides the ability to prove the
correctness and, even more usefully, test our code, this also enables
us to easily prototype a fixslicing compiler, using Coq's
\texttt{autorewrite} tactics to emulate a transformation-based
simplification engine~\citep{peyton-jones:haskell-optim} (as part of
an imaginary optimizing compiler).

@@

This article grew out of the Bachelor's research project of the first
author and the Master thesis of the second author. The former worked
on the specification, bitslicing and fixslicing of the Skinny
cipher~\citep{beierle:skinny, adomnicai:fixslicing-AES-like}. The
latter worked on the specification, bitslicing and fixslicing of the
Gift cipher~\citep{subhadeep:gift, adominicai:fixslicing-gift}. Both
ciphers are provided in the accompanying source
code\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
we shall focus exclusively on Skinny as our running example here.

Note that none of the programs below were easy to write in the first
place: it took weeks of careful study to weed out the essential
complexity from the accidental mismanagement of array indices. If the
code seem somewhat trivial, one should bear in mind that this
simplicity was hard-won. All the more reason to, some other day,
design a programming language, so as to automatically inhabit this
semantics, and write its compiler, so as to automatically optimize
code following our theorems.
